Simulated Observations of Young Gravitationally Unstable Protoplanetary Discs

### 2013-05-03 young_pp_discs.tex
neatened up figures, made first set of reviwer comments changes

\author[T.~A.~Douglas et al.]
{\parbox{\textwidth}{T. A. Douglas$^{1}$\thanks{E-mail: \texttt{pytd@leeds.ac.uk}},
P. Caselli$^{1}$,
J. D. Ilee$^{1}$,
A. C. Boley$^{2}$,
T. W. Hartquist$^{1}$,
R. H. Durisen$^{3}$, and
J. M. C. Rawlings$^{4}$\\
\vspace{0.1cm}\\
{\small{\it$^{1}$School of Physics and Astronomy, University of Leeds, Leeds LS2 9JT, UK,}} \\
{\small{\it$^{2}$Department of Astronomy, University of Florida, 211 Bryant Space Center, PO Box 112055, USA,}}\\
{\small{\it$^{3}$Department of Astronomy, Indiana University, 727 East 3rd Street, Swain West 319, Bloomington, IN 47405, USA,}}\\
{\small{\it$^{4}$Department of Physics \& Astronomy, University College London, London WC1E 6BT, UK}}\\}}

@@

\begin{abstract}
The formation and earliest stages of protoplanetary discs poorly constrained by observations. ALMA will soon revolutionise this field. Therefore, it is important to provide predictions which will be valuable for the interpretation of future high sensitivity and high angular resolution observations. Here we present simulated ALMA observations based on radiative transfer modelling of a relatively massive (0.39 M$_{\odot}$) self-gravitating disc embedded in a 10\,M$_{\odot}$ dense core, with structure similar to the pre-stellar core L1544. We focus on simple species and conclude that  C$^{17}$O 3$\rightarrow$2, HCO$^+$ 3$\rightarrow$2, OCS 26$\rightarrow$25 and H$_2$CO 4$_{04}\rightarrow$3$_{03}$ lines can be used to probe the disc structure and kinematics at all scales. 
\end{abstract}

@@

Currently few observations constrain the formation and early evolution of protoplanetary discs around solar-type and low-mass protostars, despite of the fast growing list of theoretical models on the dynamical evolution of star forming dense cores (e. g. \citealt{Krasnopolsky2011,Machida2011,Braiding2012,Joos2013}). The reason for this is that young protostars are surrounded by thick envelopes and power energetic outflows. Thus, observations of the young discs, predicted to have sizes of about 100\,au and masses as large as 10 percent the original core mass (e. g.\citealt*{Joos2012,Hayfield2011,Machida2011}), are challenging. Sensitive interferometers are needed to achieve high angular resolution and spatially/spectrally disentangle the various disc, envelope and outflow components as well as to filter out the extended emission tracing the envelope material. \smallskip

@@

To appropriately describe the environment within which a young protoplanetary disc is embedded, we combine the disc model from I2011 with a model of a dense core with characteristics similar to the well-studied pre-stellar core L1544 (\citealt{Keto2010}, hereafter KC2010). Although L1544 does not appear to contain a central protostar and disc itself, it is very similar in structure to L1521F,  another dense core in Taurus which hosts a low mass protostar and possibly a disc \citep{Bourke2006}. The KC2010 model follows the dynamical, chemical and thermal evolution of a contracting Bonnor-Ebert sphere \citep{Bonnor1956,Ebert1957} with total mass of 10\,M$_{\odot}$, until it reaches the density, temperature and velocity profiles which best match observations. The pre-stellar core model adopted here contains slight modifications due to the inclusion of oxygen cooling in the outer regions of the cloud, where CO is mostly photodissociated \citep{Caselli2012}. Fig. \ref{fig:l1544_model} shows the physical parameters of the core model. \smallskip

\begin{figure}
 \includegraphics[width=84mm]{Figures/model/L1544model_used_legend_small2.eps}
 \caption{The spherically symmetric model of the pre-stellar core L1544 used as the envelope of the young protoplanetary disc in the hybrid model. Showing gas (green) and dust (red) temperature in kelvin, log number density (blue) in cm$^{-3}$ and inward velocity / 10 (cyan) in m$\,$s$^{-1}$. Adapted from KC2010 and Keto et al. (2013, in preparation).}
 \label{fig:l1544_model}
\end{figure}

\begin{figure}
 \includegraphics[width=84mm]{Figures/model/rotational_velocities.eps}
 \caption{Azimuthaly averaged rotational velocity in the disc mid-plane. The best fit curve follows the equation$V_{rot}\frac{31.5}{\sqrt{R+1.68}}+0.518$ .}
 \label{velocity}
\end{figure}

\begin{figure*}
 \includegraphics[width=168mm]{Figures/model/rhoT5.eps}
 \caption{{\bf Left:} A 3D plot of log number density (cm$^{-3}$) showing the spiral structure in the xy plane and scale height of the disc. {\bf Right:} The 3D temperature (K) structure of the disc; regions cooler than 40$\,$K are not shown in 3D, in order to highlight the narrow central region containing hot material.}
 \label{rhoT} 
\end{figure*}

\begin{figure*}
 \includegraphics[width=168mm]{Figures/model/columnDensities3.eps}
 \caption{Column density distributions of OCS, H$_2$CO, HCO$^+$ and CO across the disc, as calculated by I2011. Note the different zones traced by the various species, with OCS mostly tracing the inner regions and HCO$^+$ probing the gas at larger radii.}
 \label{Chemistry} 
\end{figure*}

The pre-stellar core structure is maintained down to a radius of 80\,au, within which the model of the young protoplanetary disc is used. The disc structure used is from the same hydrodynamics simulation described in I2011, which is a 1$\,$M$_\odot$ protostar surrounded by a 0.39$\,$M$_\odot$ disc. Fig.~\ref{velocity} shows the azimuthally averaged rotational velocity in the disc plane, together with the best fit curve, which will be used in Sec.~\ref{sec:alma_predictions} to compare with position velocity diagrams. The system is envisaged to be at a very early stage of evolution and embedded in an envelope.  The still growing protostar will eventually become an A star through accretion of most of the disc's mass. The hydrodynamics simulation is run with a proper equation of state for molecular hydrogen \citep{Boley2007}.  The star is allowed to move freely in response to disc torques, and radiative cooling is included \citep{Boley2009}. The self-gravitating disc exhibits prominent spiral structure, with H$_2$ number densities in the disc ranging from 10$^4$-10$^{13}$ cm$^{-3}$ and temperatures from 30-400 K (fig. \ref{rhoT}). For the non-LTE 3D radiative transfer modelling presented here, we interpolate the I2011 simulation output on to a 256$\times$256$\times$64 Cartesian grid with spatial resolution of 0.5$\,$au in x, y and z and sample from this to create the unstructured radiative transfer grid (see section \ref{subsec:gridding}).  As in the hydrodynamics simulation, the dust and gas temperatures are assumed to be in equilibrium within the disc and a gas to dust mass ratio of 100 is used throughout the entirety of the model. The dust opacities were adopted from \citet{Ossenkopf1994} and refer to dust grains with thick icy mantles and 10$^6$ yr coagulation history.  Although dust grains within the protoplanetary disc are expected to coagulate further,  significantly affecting the opacity, we did not take grain growth into account in our model.  The dust opacities used in the radiative transfer are not the same as the ones used in the hydrodynamic model (which used the dust opacities of \citealt{DAlessio2001}). However, the results described here do not depend on the details of the dust opacity used during the hydrodynamic evolution.\smallskip

\subsection{Chemical structure} \label{subsec:chemical_structure}
Chemical abundances in the disc were taken from I2011, who followed gas-grain chemical processes during the dynamical evolution of the disc. As described in I2011, the hydrodynamical simulation included Lagrangian tracer fluid elements in addition to the solution of the equations of hydrodynamics on an Eulerian grid.  These fluid elements were used to record the thermal history of the gas as material passed through and between the spiral structure. I2011 used this time evolution information from fluid elements to calculate the abundances of 125 species, related by 1334 reactions.  These abundances were then interpolated by I2011 on to a Cartesian grid with cell sizes 2.2$\times$2.2$\times$0.22$\,$au$^3$ in x, y and z to reveal the morphology of the disc as seen by different molecular species. From these we selected the four species which appear to trace different regions of the disc: the inner 30\,au (OCS), the inner 40\,au (H$_2$CO),  the region between $\simeq$40 and 60\,au (HCO$^+$) and the entirety of the disc (C$^{17}$O) (see fig. \ref{Chemistry}). The abundance model used for C$^{17}$O was the CO model reduced by a factor of 1792, the ratio of $^{16}$O to $^{17}$O in the local ISM \citep{Wilson1994}. As explained by I2011,  H$_2$CO and OCS mostly probe the central warm regions, where icy mantles evaporate, whereas HCO$^+$ preferentially traces the outer spiral pattern as in the central region it is destroyed by water molecules and transformed into H$_3$O$^+$ and CO. The simple chemistry in the KC2010 model, adopted here as the envelope of the protoplanetary disc, does not provide detailed abundances of molecular species (besides CO and H$_2$O, see also \citealt{Caselli2012}). As discussed in the results section \ref{sec:model_results}, abundances have been adopted based on values measured toward similar objects. The molecular data used here for the radiative transfer are taken from the Leiden Atomic and Molecular DAtabase (LAMDA) (\citealt{Schoier2005} http://home.strw.leidenuniv.nl/$\sim$moldata/).

@@

\begin{figure}
 \includegraphics[width=84mm]{Figures/model/lime3.eps}
 \caption{A plot of the points selected by the gridding process and the paths down which photons can propagate for points in the central r, $\theta$ plane. The points are colour coded by the density distribution (in m$^{-3}$, as used in LIME) and are more concentrated at small radii and in the most dense regions. The circular high density region in the centre is the disc model and is 128$\,$au in diameter.}
 \label{grid}
\end{figure}

@@

\begin{figure}
 \includegraphics[width=84mm]{Figures/sim/continuum2.eps}
 \caption{{\bf Left:} A 300$\,$GHz continuum image of the model, created from the average of ten LIME runs. {\bf Right:} the output of a single radiative transfer simulation, with artefacts due to finite gridding.}
 \label{averages}
\end{figure}

\begin{figure}
 \includegraphics[width=84mm]{Figures/model/lime_points_rz_histo3.eps}
 \caption{A 2D histogram of the point distribution throughout the model. The disc and envelope can be seen as two separate entities which have to be sampled using different point distributions.}
 \label{points}
\end{figure}

@@

\begin{figure*}
 \includegraphics[width=150mm]{Figures/sim/imageALL_30deg_all2.eps}
 \caption{{\bf Left:} Continuum subtracted integrated intensity maps. {\bf Right:} Intensity weighted velocity maps. The spiral structures in different regions of the disc are highlighted by different molecular lines. From top to bottom the lines displayed are: OCS 26$\rightarrow$25, H$_2$CO 4$_{04}$$\rightarrow$3$_{03}$, C$^{17}$O 3$\rightarrow$2 and HCO$^+$ 3$\rightarrow$2. All maps are obtained from integration over -12.5 to -0.5 and 0.5 to 12.5 km$\,$s$^{-1}$ in order to avoid the envelope contribution.}
 \label{sim_all}
\end{figure*}

@@

\begin{figure*}
 \includegraphics[width=110mm]{Figures/sim/appendix_inclinations2.eps}
 \caption{C$^{17}$O 3$\rightarrow$2 {\bf left:} Continuum subtracted integrated intensity map (in K$\,$km$\,$s$^{-1}$). {\bf Centre:} Intensity weighted velocity map (in km$\,$s$^{-1}$). {\bf Right:} Position-velocity diagram with intensities in K. The inclinations for which results are shown from top to bottom are: 75$^\circ$, 45$^\circ$, 30$^\circ$ and 15$^\circ$ to edge on.}
 \label{other_inc}
\end{figure*}

@@

In order to make predictions of the observability of the features for which we have constructed synthetic brightness maps, the Common Astronomy Software Applications package (CASA) was used to simulate observations with the completed ALMA in the 26th most extended configuration (out of a total of 28). The longest baseline of this configuration is 14.4\,km and the beam sizes range between 0.02 and 0.03 arcseconds at the selected frequencies. This angular resolution is close to the size of the spiral structure in the disc. As the weakest lines studied in the model are of the order of 0.1-0.2$\,$mJy$\,$beam$^{-1}$ at this resolution, Band 7 sensitivities of the order of 0.02$\,$mJy$\,$beam$^{-1}$ are required. This implies around 6 hours of integration time. The exact sensitivities are given in Table \ref{sigmas}.
\begin{table*}
  \centering
  \begin{minipage}{90mm}
    \caption{Line sensitivities obtained with ALMA after 6 hours integration time with velocity resolution of 400$\,$m\,s$^{-1}$, obtained with the ALMA on-line sensitivity calculator.}
    \label{sigmas}
    \begin{tabular}{c||c|c|c|c}
      \hline
      Species & Transition & Frequency & Upper energy level & Sensitivity\\
                      &                    & (GHz)            & (K)                            & (mJy beam$^{-1}$) \\
      \hline
      C$^{17}$O & 3$\rightarrow$2 & 337.06 & 32.3 & 0.0270 \\
      OCS & 26$\rightarrow$25 & 316.15 & 205 & 0.0275 \\
      HCO$^+$ & 3$\rightarrow$2 & 267.56 & 25.7 & 0.0176 \\
      H$_2$CO & 4$_{04}\rightarrow$3$_{03}$ &  290.62 & 34.9 & 0.0228 \\
      \hline
    \end{tabular}
  \end{minipage}
\end{table*}
In order to clean the images which featured mainly absorption, the sky model used was the output of the LIME simulations with the continuum subtracted and then the spectrum inverted. The simulated images were cleaned down to a 3$\sigma$ level with $\sigma$ (the sensitivity) given in Table \ref{sigmas}.\smallskip

\begin{figure}
 \includegraphics[width=84mm]{Figures/sim/casa_cont_300GHz_invert2.eps}

@@

\begin{figure*}
 \includegraphics[width=168mm]{Figures/sim/channel_map-2.eps} 
 \caption{H$_2$CO 4$_{04}\rightarrow$3$_{03}$ channel maps as simulated with CASA. The contours start at 0.35 mJy$\,$beam$^{-1}\,$km$\,$s$^{-1}$  and are in steps of 0.35 mJy$\,$beam$^{-1}\,$km$\,$s$^{-1}$ (absorption in blue, emission in red). Overlaid on the 1mm continuum emission. the integration range for each map is 2.4 km$\,$s$^{-1}$, for the top left the range is from -10.8 to -8.4 km$\,$s$^{-1}$ and for the bottom right the range is from 8.4 to 10.8 km$\,$s$^{-1}$}
 \label{h2co_chanmap}
\end{figure*}

@@

\begin{figure}
 \includegraphics[width=84mm]{Figures/sim/casa_C17O_15deg_all3.eps}

@@

\begin{figure*}
 \includegraphics[width=140mm]{Figures/sim/casa_all_30deg_contSub_dots.eps}
 \caption{Integrated intensity maps for the simulated observations of: {\bf Top Left:} H$_2$CO 4$_{04}\rightarrow$3$_{03}$, {\bf Top Right:} C$^{17}$O 3$\rightarrow$2, {\bf Bottom Left:} OCS 26$\rightarrow$25 and {\bf Bottom Right:} HCO$^+$ 3$\rightarrow$2. Contours start at 3$\sigma$ and increase in intervals of 3$\sigma$ for each of the maps except the H$_2$CO contours, which start at the 5$\sigma$ level and increase in intervals of 5$\sigma$. Emission is in red contours, absorption in blue. The coloured dots refer to the positions from which the spectra shown in fig. \ref{spectra} are taken.}
\label{mom0_maps}
\end{figure*}

@@

\begin{figure*}
 \includegraphics[width=180mm]{Figures/sim/casa_all_30deg_PV_rotCurve_small2.eps}
 \caption{Position velocity diagrams for each of the lines simulated with CASA. The green curve on each diagram is the best fit curve to the azimuthally averaged rotation profile from fig. 4.}
 \label{pvs}
\end{figure*}

@@

\begin{figure*}
 \includegraphics[width=168mm]{Figures/sim/casa_all_spectra2.eps}
 \caption{Simulated spectra for each of the selected transitions at the positions indicated by the coloured dots in fig. \ref{mom0_maps} (the pink, green and cyan curves are taken from the locations marked by the dots of the same colour). In each diagram, the three spectra are taken from three positions on the Y=0 axis, one is at X=0 (green) and the other 2 are from x=$\pm\,$N (plus yellow, minus pink) where N is the distance from the centre: 10 au for OCS, 20 au for H$_2$CO, 28 au for C$^{17}$O, 38 au for HCO$^+$.}
 \label{spectra}
\end{figure*}

@@

These results demonstrate that ALMA will be able to probe the very earliest stages of circumstellar discs which feed the growing protostar and eventually will evolve in a planetary system. Our results show that ALMA observations in band 7 will allow the assessment of whether or not these young discs are gravitationally unstable and test theories of pre-stellar $\rightarrow$ protostellar core evolution. If such gravitationally instabilities exist they will have significant implications for the evolution of the disc. For example the will affect fragmentation, planetesimal formation \citep{Boley2009,Johnson2013,Gibbons2012} and the chemical evolution in the disc (I2011).

@@

\section*{Acknowledgements}
PC and TWH acknowledge support of successive rolling grants awarded by the UK Science and Technology Funding Council. 
TAD and JDI acknowledge studentships from the Science and Technology Facilities Council of the United Kingdom (STFC).
The research leading to these results has received funding from the European Union FP7-2011 under grant agreement no 284405.
TAD would like to thank Nicol Cunningham for her assistance with CASA and Christian Brinch for his radiative transfer code LIME and for guidance on using it.

>>>>

\author[T.~A.~Douglas et al.]
{\parbox{\textwidth}{T. A. Douglas$^{1}$\thanks{E-mail: \texttt{pytd@leeds.ac.uk}},
P. Caselli$^{1}$,
J. D. Ilee$^{5,1}$,
A. C. Boley$^{2}$,
T. W. Hartquist$^{1}$,
R. H. Durisen$^{3}$, and
J. M. C. Rawlings$^{4}$\\
\vspace{0.1cm}\\
{\small{\it$^{1}$School of Physics and Astronomy, University of Leeds, Leeds LS2 9JT, UK,}} \\
{\small{\it$^{2}$Department of Astronomy, University of Florida, 211 Bryant Space Center, PO Box 112055, USA,}}\\
{\small{\it$^{3}$Department of Astronomy, Indiana University, 727 East 3rd Street, Swain West 319, Bloomington, IN 47405, USA,}}\\
{\small{\it$^{4}$Department of Physics \& Astronomy, University College London, London WC1E 6BT, UK}}\\
{\small{\it$^{5}$School of Physics and Astronomy, University of St Andrews, North Haugh, St Andrews, KY16 9SS, UK,}}\\
}}

@@

\begin{abstract}
The formation and earliest stages of protoplanetary discs remain poorly constrained by observations. ALMA will soon revolutionise this field. Therefore, it is important to provide predictions which will be valuable for the interpretation of future high sensitivity and high angular resolution observations. Here we present simulated ALMA observations based on radiative transfer modelling of a relatively massive (0.39 M$_{\odot}$) self-gravitating disc embedded in a 10\,M$_{\odot}$ dense core, with structure similar to the pre-stellar core L1544. We focus on simple species and conclude that  C$^{17}$O 3$\rightarrow$2, HCO$^+$ 3$\rightarrow$2, OCS 26$\rightarrow$25 and H$_2$CO 4$_{04}\rightarrow$3$_{03}$ lines can be used to probe the disc structure and kinematics at all scales. 
\end{abstract}

@@

Currently few observations constrain the formation and early evolution of protoplanetary discs around solar-type and low-mass protostars, despite the fast growing list of theoretical models on the dynamical evolution of star forming dense cores (e. g. \citealt{Krasnopolsky2011,Machida2011,Braiding2012,Joos2013}). The reason for this is that young protostars are surrounded by thick envelopes and power energetic outflows. Thus, observations of the young discs, predicted to have sizes of about 100\,au and masses as large as 10 percent the original core mass (e. g. \citealt{Joos2012,Hayfield2011,Machida2011}), are challenging. Sensitive interferometers are needed to achieve high angular resolution and spatially/spectrally disentangle the various disc, envelope and outflow components as well as to filter out the extended emission tracing the envelope material. \smallskip

@@

To appropriately describe the environment within which a young protoplanetary disc is embedded, we combine the disc model from I2011 with a model of a dense core with characteristics similar to the well-studied pre-stellar core L1544 (\citealt{Keto2010}, hereafter KC2010). Although L1544 does not appear to contain a central protostar and disc itself, it is very similar in structure to L1521F,  another dense core in Taurus which hosts a low mass protostar and possibly a disc \citep{Bourke2006}. The KC2010 model follows the dynamical, chemical and thermal evolution of a contracting Bonnor-Ebert sphere \citep{Bonnor1956,Ebert1957} with total mass of 10\,M$_{\odot}$, until it reaches the density, temperature and velocity profiles which best match observations. The pre-stellar core model adopted here contains slight modifications due to the inclusion of oxygen cooling in the outer regions of the cloud, where CO is mostly photodissociated \citep{Caselli2012}. Fig. \ref{fig:l1544_model} shows the physical parameters of the core model{\bf as used in an improved version of the KC2010 model (Keto et al. 2013, in preparation).}\smallskip

\begin{figure}
 \includegraphics[width=84mm]{Figures/model/fig01.eps}
 \caption{The spherically symmetric model of the pre-stellar core L1544 used as the envelope of the young protoplanetary disc in the hybrid model. Showing gas (green) and dust (red) temperature in kelvin, log number density (blue) in cm$^{-3}$ and inward velocity / 10 (cyan) in m$\,$s$^{-1}$. Adapted from KC2010 and Keto et al. (2013, in preparation).}
 \label{fig:l1544_model}
\end{figure}

\begin{figure}
 \includegraphics[width=84mm]{Figures/model/fig02.eps}
 \caption{Azimuthaly averaged rotational velocity in the disc mid-plane. The best fit curve follows the equation$V_{\rm rot}=\frac{31.5}{\sqrt{R+1.68}}+0.518$ .}
 \label{velocity}
\end{figure}

\begin{figure*}
 \includegraphics[width=168mm]{Figures/model/fig03.eps}
 \caption{{\bf Left:} A 3D plot of log number density (cm$^{-3}$) showing the spiral structure in the xy plane and scale height of the disc. {\bf Right:} The 3D temperature (K) structure of the disc; regions cooler than 40$\,$K are not shown in 3D, in order to highlight the narrow central region containing hot material.}
 \label{rhoT} 
\end{figure*}

\begin{figure*}
 \includegraphics[width=168mm]{Figures/model/fig04.eps}
 \caption{Column density distributions of OCS, H$_2$CO, HCO$^+$ and CO across the disc, as calculated by I2011. Note the different zones traced by the various species, with OCS mostly tracing the inner regions and HCO$^+$ probing the gas at larger radii.}
 \label{Chemistry} 
\end{figure*}

The pre-stellar core structure is maintained down to a radius of 80\,au, within which the model of the young protoplanetary disc is used. The disc structure used is from the same hydrodynamics simulation described in I2011, which is a 1$\,$M$_\odot$ protostar surrounded by a 0.39$\,$M$_\odot$ disc. Fig.\ref{velocity} shows the azimuthally averaged rotational velocity in the disc plane, together with the best fit curve, which will be used in Sec.~\ref{sec:alma_predictions} to compare with position velocity diagrams. The system is envisaged to be at a very early stage of evolution and embedded in an envelope.  The still growing protostar will eventually become an A star through accretion of most of the disc's mass. The hydrodynamics simulation is run with a proper equation of state for molecular hydrogen \citep{Boley2007}.  The star is allowed to move freely in response to disc torques, and radiative cooling is included \citep{Boley2009}. The self-gravitating disc exhibits prominent spiral structure, with H$_2$ number densities in the disc ranging from 10$^4$-10$^{13}$ cm$^{-3}$ and temperatures from 30-400 K (fig. \ref{rhoT}). For the non-LTE 3D radiative transfer modelling presented here, we interpolate the I2011 simulation output on to a 256$\times$256$\times$64 Cartesian grid with spatial resolution of 0.5$\,$au in x, y and z and sample from this to create the unstructured radiative transfer grid (see section \ref{subsec:gridding}).  As in the hydrodynamics simulation, the dust and gas temperatures are assumed to be in equilibrium within the disc and a gas to dust mass ratio of 100 is used throughout the entirety of the model. The dust opacities were adopted from \citet{Ossenkopf1994} and refer to dust grains with thick icy mantles and 10$^6$ yr coagulation history.  Although dust grains within the protoplanetary disc are expected to coagulate further,  significantly affecting the opacity, we did not take grain growth into account in our model.  The dust opacities used in the radiative transfer are not the same as the ones used in the hydrodynamic model (which used the dust opacities of \citealt{DAlessio2001}). However, the results described here do not depend on the details of the dust opacity used during the hydrodynamic evolution.\smallskip

\subsection{Chemical structure} \label{subsec:chemical_structure}
Chemical abundances in the disc were taken from I2011, who followed gas-grain chemical processes during the dynamical evolution of the disc. As described in I2011, the hydrodynamical simulation included Lagrangian tracer fluid elements in addition to the solution of the equations of hydrodynamics on an Eulerian grid.  These fluid elements were used to record the thermal history of the gas as material passed through and between the spiral structure. I2011 used this time evolution information from fluid elements to calculate the abundances of 125 species, related by 1334 reactions.  These abundances were then interpolated by I2011 on to a Cartesian grid with cell sizes 2.2$\times$2.2$\times$0.22$\,$au$^3$ in x, y and z to reveal the morphology of the disc as seen by different molecular species. From these we selected the four species which appear to trace different regions of the disc: the inner 30\,au (OCS), the inner 40\,au (H$_2$CO),  the region between $\simeq$40 and 60\,au (HCO$^+$) and the entirety of the disc (C$^{17}$O) (see fig. \ref{Chemistry}). The abundance model used for C$^{17}$O was the CO model reduced by a factor of 1792, the ratio of $^{16}$O to $^{17}$O in the local ISM \citep{Wilson1994}. As explained by I2011,  H$_2$CO and OCS mostly probe the central warm regions, where icy mantles evaporate, whereas HCO$^+$ preferentially traces the outer spiral pattern as in the central region it is destroyed by water molecules and transformed into H$_3$O$^+$ and CO. The simple chemistry in the KC2010 model, adopted here as the envelope of the protoplanetary disc, does not provide detailed abundances of molecular species (besides CO and H$_2$O, see also \citealt{Caselli2012}). As discussed in the results section \ref{sec:model_results}, abundances have been adopted based on values measured toward similar objects. The molecular data used here for the radiative transfer are taken from the Leiden Atomic and Molecular DAtabase (LAMDA) (\citealt{Schoier2005} http://home.strw.leidenuniv.nl/$\sim$moldata/). \smallskip

{\bf Whilest photochemistry was not included in I2011, it is unclear to what extent the radiation from the protostar will impact upon the disc, given that material infalling from the envelope and newly forming outflows will be obscuring the star. Column densities between upper layers of the disc and the protostar in the model used here are on the order of 10$^{23}$ cm$^{-2}$. This is in contrast to later stages of disc evolution where the envelope has been almost entirely cleared (e.g. Walsh et. al), in this case it is clear that the chemistry in the upper layers of the disc will be dominated by photo-chemical processes.}

@@

\begin{figure}
 \includegraphics[width=84mm]{Figures/model/fig05.eps}
 \caption{A plot of the points selected by the gridding process and the paths down which photons can propagate for points in the central r, $\theta$ plane. The points are colour coded by the density distribution (in m$^{-3}$, as used in LIME) and are more concentrated at small radii and in the most dense regions. The circular high density region in the centre is the disc model and is 128$\,$au in diameter.}
 \label{grid}
\end{figure}

@@

\begin{figure}
 \includegraphics[width=84mm]{Figures/sim/fig06.eps}
 \caption{{\bf Left:} A 300$\,$GHz continuum image of the model, created from the average of ten LIME runs. {\bf Right:} the output of a single radiative transfer simulation, with artefacts due to finite gridding.}
 \label{averages}
\end{figure}

\begin{figure}
 \includegraphics[width=84mm]{Figures/model/fig07.eps}
 \caption{A 2D histogram of the point distribution throughout the model. The disc and envelope can be seen as two separate entities which have to be sampled using different point distributions.}
 \label{points}
\end{figure}

@@

\begin{figure*}
 \includegraphics[width=150mm]{Figures/sim/fig08.eps}
 \caption{{\bf Left:} Continuum subtracted integrated intensity maps. {\bf Right:} Intensity weighted velocity maps. The spiral structures in different regions of the disc are highlighted by different molecular lines. From top to bottom the lines displayed are: OCS 26$\rightarrow$25, H$_2$CO 4$_{04}$$\rightarrow$3$_{03}$, C$^{17}$O 3$\rightarrow$2 and HCO$^+$ 3$\rightarrow$2. All maps are obtained from integration over -12.5 to -0.5 and 0.5 to 12.5 km$\,$s$^{-1}$ in order to avoid the envelope contribution.}
 \label{sim_all}
\end{figure*}

@@

\begin{figure*}
 \includegraphics[width=110mm]{Figures/sim/fig09.eps}
 \caption{C$^{17}$O 3$\rightarrow$2 {\bf left:} Continuum subtracted integrated intensity map (in K$\,$km$\,$s$^{-1}$). {\bf Centre:} Intensity weighted velocity map (in km$\,$s$^{-1}$). {\bf Right:} Position-velocity diagram with intensities in K. The inclinations for which results are shown from top to bottom are: 75$^\circ$, 45$^\circ$, 30$^\circ$ and 15$^\circ$ to edge on.}
 \label{other_inc}
\end{figure*}

@@

In order to make predictions of the observability of the features for which we have constructed synthetic brightness maps, the Common Astronomy Software Applications package (CASA) was used to simulate observations with the completed ALMA in the 26th most extended configuration (out of a total of 28). The longest baseline of this configuration is 14.4\,km and the beam sizes range between 0.02 and 0.03 arcseconds at the selected frequencies. This angular resolution is close to the size of the spiral structure in the disc. As the weakest lines studied in the model are of the order of 0.1-0.2$\,$mJy$\,$beam$^{-1}$ at this resolution, Band 7 sensitivities of the order of 0.02$\,$mJy$\,$beam$^{-1}$ are required. This implies around 6 hours of integration time. The exact sensitivities are given in Table \ref{sigmas}.
\begin{table*}
  \centering
  \begin{minipage}{90mm}
    \caption{Line sensitivities obtained with ALMA after 6 hours integration time with velocity resolution of 400$\,$m\,s$^{-1}$, obtained with the ALMA on-line sensitivity calculator.}
    \label{sigmas}
    \begin{tabular}{c||c|c|c|c}
      \hline
      Species & Transition & Frequency & Upper energy level & Sensitivity\\
              &            &   (GHz)   &       (K)          & (mJy beam$^{-1}$) \\
      \hline
      C$^{17}$O & 3$\rightarrow$2 & 337.06 & 32.3 & 0.0270 \\
      OCS & 26$\rightarrow$25 & 316.15 & 205 & 0.0275 \\
      H$_2$CO & 4$_{04}\rightarrow$3$_{03}$ &  290.62 & 34.9 & 0.0228 \\
      HCO$^+$ & 3$\rightarrow$2 & 267.56 & 25.7 & 0.0176 \\
      \hline
    \end{tabular}
  \end{minipage}
\end{table*}
In order to clean the images which featured mainly absorption, the sky model used was the output of the LIME simulations with the continuum subtracted and then the spectrum inverted. The simulated images were cleaned down to a 3$\sigma$ level with $\sigma$ (the sensitivity) given in Table \ref{sigmas}.\smallskip

\begin{figure}
 \includegraphics[width=84mm]{Figures/sim/fig10.eps}

@@

\begin{figure*}
 \includegraphics[width=168mm]{Figures/sim/fig11.eps} 
 \caption{H$_2$CO 4$_{04}\rightarrow$3$_{03}$ channel maps as simulated with CASA. The contours start at 0.35 mJy$\,$beam$^{-1}\,$km$\,$s$^{-1}$  and are in steps of 0.35 mJy$\,$beam$^{-1}\,$km$\,$s$^{-1}$ (absorption in blue, emission in red). Overlaid on the 1mm continuum emission. the integration range for each map is 2.4 km$\,$s$^{-1}$, for the top left the range is from -10.8 to -8.4 km$\,$s$^{-1}$ and for the bottom right the range is from 8.4 to 10.8 km$\,$s$^{-1}$}
 \label{h2co_chanmap}
\end{figure*}

@@

\begin{figure}
 \includegraphics[width=84mm]{Figures/sim/fig12.eps}

@@

\begin{figure*}
 \includegraphics[width=140mm]{Figures/sim/fig13.eps}
 \caption{Integrated intensity maps for the simulated observations of: {\bf Top Left:} H$_2$CO 4$_{04}\rightarrow$3$_{03}$, {\bf Top Right:} C$^{17}$O 3$\rightarrow$2, {\bf Bottom Left:} OCS 26$\rightarrow$25 and {\bf Bottom Right:} HCO$^+$ 3$\rightarrow$2. Contours start at 3$\sigma$ and increase in intervals of 3$\sigma$ for each of the maps except the H$_2$CO contours, which start at the 5$\sigma$ level and increase in intervals of 5$\sigma$. Emission is in red contours, absorption in blue. The coloured dots refer to the positions from which the spectra shown in fig. \ref{spectra} are taken.}
\label{mom0_maps}
\end{figure*}

@@

\begin{figure*}
 \includegraphics[width=180mm]{Figures/sim/fig14.eps}
 \caption{Position velocity diagrams for each of the lines simulated with CASA. The green curve on each diagram is the best fit curve to the azimuthally averaged rotation profile from fig. 4.}
 \label{pvs}
\end{figure*}

@@

\begin{figure*}
 \includegraphics[width=168mm]{Figures/sim/fig15.eps}
 \caption{Simulated spectra for each of the selected transitions at the positions indicated by the coloured dots in fig. \ref{mom0_maps} (the pink, green and cyan curves are taken from the locations marked by the dots of the same colour). In each diagram, the three spectra are taken from three positions on the Y=0 axis, one is at X=0 (green) and the other 2 are from x=$\pm\,$N (plus yellow, minus pink) where N is the distance from the centre: 10 au for OCS, 20 au for H$_2$CO, 28 au for C$^{17}$O, 38 au for HCO$^+$.}
 \label{spectra}
\end{figure*}

@@

{\bf The disc model used does not include magnetic fields, including magnetic fields may suppress GI and spiral wave formation by slowing down or even preventing early disc growth in the case of sufficicent magnetization and alinegment between the parent cores rotation axis and maxnetic field direction (Joos et al. 2012, Li et al. 2013). This would mean that if observations similar to the ones we have descirbed are made constraints could be placed on B-field strength within cores/discs. \smallskip Future work including looking at the effects of material from the envelope falling onto the disc and how this affects the chemistry and line propagation, and including realistic outflows into the model.}

These results demonstrate that ALMA will be able to probe the very earliest stages of circumstellar discs which feed the growing protostar and eventually will evolve in a planetary system. Our results show that ALMA observations in band 7 will allow the assessment of whether or not these young discs are gravitationally unstable and test theories of pre-stellar to protostellar core evolution. If such gravitational instabilities exist they will have significant implications for the evolution of the disc. For example the will affect fragmentation, planetesimal formation \citep{Boley2009,Johnson2013,Gibbons2012} and the chemical evolution in the disc (I2011).

@@

\section*{Acknowledgements}
PC and TWH acknowledge support of successive rolling grants awarded by the UK Science and Technology Funding Council.  
TAD and JDI acknowledge studentships from the Science and Technology Facilities Council of the United Kingdom (STFC), JDI also acknowledges funding from the European Union FP7-2011 under grant agreement no 284405. 
TAD would like to thank Nichol Cunningham for her assistance with CASA and Christian Brinch for his radiative transfer code LIME and for guidance on using it.
